---
title: "Practical Machine Learning project"
output: html_document
---

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset). 

### Data 
The training data for this project are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.  

```{r, results = FALSE, cache = TRUE}
require(caret)
require(foreach)
require(parallel)
require(data.table)
set.seed(100)
setwd("~/git/ML-Project")
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
download.file(url, destfile="pml-training.csv", method = 'curl')
dataTrain <- read.csv("pml-training.csv")
```

The objective is to predict the classe variable

```{r, results = TRUE, cache = TRUE}
str(dataTrain$classe)
summary(dataTrain$classe)
```

```{r, results = TRUE, echo=FALSE, cache = TRUE}
plot(dataTrain$classe, main = "Classe distribution")
```

We will use variables with complete rows, and reduce the dataset. We will also remove Name, ID, and timestamp variables that may interfere and not help the model
```{r, results = FALSE, cache = TRUE}
varIncomplete <- sapply(dataTrain, function (x) any(is.na(x) | x == ""))
dataTrainPrediction <- dataTrain[, !varIncomplete]
dataTrainPrediction <- dataTrainPrediction[, !grepl("X|user_name|.timestamp.|.timestamp|.window", names(dataTrainPrediction))]
```

```{r, results = TRUE, cache = TRUE}
dim(dataTrainPrediction)
```

Next, we will partition and pre-process the training dataset
```{r, results = TRUE, cache = TRUE}
# Partitioning
part <- createDataPartition(y=dataTrainPrediction$classe, p=0.6, list=FALSE)
dataTrainModel <- dataTrainPrediction[part, ]
dataProbeModel <- dataTrainPrediction[-part, ]
preProc <- preProcess(dataTrainModel)
preProc
dataTrainModelPC <- predict(preProc, dataTrainModel) # Apply to train set
preProc <- preProcess(dataProbeModel)
dataProbeModelPC <- predict(preProc, dataProbeModel) # Apply to Probe set
```

## Select a model and train
The dataset looks suitable for the RF (random forest) prediction model. The error threshold we will use to measure success is 3%. We will use the parallel processing RF methid within caret.

```{r, results = TRUE, cache = TRUE}
classeModel <- train(classe ~ ., data=dataTrainModelPC, method="parRF")
classeModel
```

We now try the model against the probing dataset

```{r, results = TRUE, cache = TRUE}
predProbe <- predict(classeModel, dataProbeModelPC)
confusionMatrix(predProbe, dataProbeModelPC$classe)
classeModel$finalModel
```

The model seems to have 1% error rate (great!)

A plot of the final model:
```{r, results = TRUE, echo = FALSE, cache = TRUE}
plot(varImp(classeModel), main = "Final model variable weights")
```

```{r, results = FALSE, echo = FALSE, cache = TRUE}
save (classeModel, file="classeModel.RData") # we save the model for later use
```

## Use the model to predict Testing set
We load and prepare the data
```{r, results = FALSE, cache = TRUE}
load(file="classeModel.RData", verbose=TRUE) # re-loading the model
url2 <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
download.file(url2, destfile="pml-testing.csv", method = 'curl')
dataTest <- read.csv("pml-testing.csv")
#preProc <- preProcess(dataTest)
dataTestPC <- predict(preProc, dataTest) # Pre-process test dataset
```

Predict:
```{r, results = TRUE, cache = TRUE}
pred_classe <- predict(classeModel, dataTestPC)
dataTestPC <- cbind(pred_classe, dataTestPC)
dataTestPC[, c("user_name","pred_classe")]
```

```{r, results = TRUE, echo=FALSE, cache = TRUE}
plot(dataTestPC$pred_classe, main = "Classe prediction distribution")
```

```{r, results = FALSE, echo=FALSE, cache = TRUE}
# Coursera files:
pml_write_files = function(x){
  n = length(x)
  path <- "answers"
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=file.path(path, filename),quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(pred_classe)
```

